### 2018-01-27 gf-pldi-2018/RESPONSE-PLDI-2018.txt

Response Part 1: clarifications & corrections
================================================================================

#99A says

> This paper particularly considers five different embeddings, including the
> erasure embedding ... and locally-defensive embedding, which removes many
> unnecessary runtime checks from natural embedding

The locally-defensive embedding compiles the source so that the execution 
skips checks that are **necessary** to enforce _type-soundness_ and
unnecessary for the weaker notion of _tag-soundness_.

@@

Pycket is an orthogonal result
--------------------------------------------------------------------------------

#99A says

> While the locally-defensive embedding in this paper improves the performance
> of the natural embedding by about a magnitude, the recent work by Pycket has
> already achieved this.

The Pycket work demonstrates that replacing a conventional JIT with a tracing
JIT can significantly reduce the overhead of natural-embedding type soundness
(one and the same semantics).

Our work demonstrates that replacing type soundness with type-tag soundness
can lead to significant performance improvements (different semantics).

@@

These are separate, orthogonal results.

Combining these two independent pieces of work is future work and will
require close collaboration with the independent Pycket team.

@@

#99C says

> There is perhaps another contribution: the idea of "locally-defensive"
> embedding.  Frustratingly, I can't figure out if this is an original
> contribution of the paper, or if it's taken from somewhere else.

The locally-defensive embedding is a mathematical characterization of
transient semantics, and our locally-defensive Racket implements it.

In contrast, the POPL 2017 paper by Vitousek et al. defines transient by
its implementation, which does not explain the properties that an
implementation needs to satisfy in order to be called "transient". So it is
unclear, for example, how to give a transient semantics for the
gradually-typed lambda calculus or a language that is not Python. Our work
fills this gap.

Vitousek et al. [POPL 2017] does not compare type soundness and tag soundness
--------------------------------------------------------------------------------

@@

#99B says

> The main result is a confirmation that the transient semantics of Vitousek
> (POPL 2017) is indeed better-performing than the usual semantics with eager
> casts. This is hardly surprising given the experiments in that paper.

Review #99B sells our results short:

1. Vitousek et al. do _not_ report the performance of programs that mix typed
   and untyped code. (They report the performance of fully typed Reticulated
   programs vs. Python programs.)

@@

2. Our paper provides the first controlled comparison of type soundness and
   type-tag soundness on the _same_ language platform on mixed-typed code.

3. Our paper provides a mathematical comparison between type soundness and tag
   soundness (Vitousek et al. conflate the two; see their Corollary 5.5.1).

Our paper aims to present scientific novelties, not surprising ones.

@@

The idea of a spectrum has NOT been previously explored
--------------------------------------------------------------------------------

#99B says:

> The other claimed contribution is the idea that there is in fact a spectrum
> of soundness and performance: but this has been previously explored in the
> work on like types, at least.

Review #99B mischaracterizes like types. The like types approach describes a
_novel type system_ with a **conventional** notion of type soundness. (BTW,
it is also restricted to first--order settings.)

Our paper is the first to show how _one language_ can have _multiple semantics_,
each guaranteeing a _different degree of soundness_.

@@

Sound gradual typing is a worthwhile goal
--------------------------------------------------------------------------------

#99C says

> There is a big literature on gradual typing, but only one form appears to have
> had practical impact.  That's the one that is here called the "erasure
> embedding".  The obvious question is: are the other approaches worth pursuing
> at all?

Industry chose "erasure" (optional typing) to avoid obvious performance
penalties. But, this is an unsound form of gradual typing.  Experience, and
our theorem (L_E term safety), show that this choice lets programs produce
wrong answers due to uncaught violations of the static typing discipline.
This has two negative consequences: users may have to cope with subtly
wrong outputs whose faults may affect real life long after running the
program (say lost elections, see SNAPL '17), and programmers cannot use
types to reason about correctness (see theorems).

We proudly continue to investigate sound gradual typing because we are
academic researchers who do not wish to give up after a mere decade to get
things right and because sound gradual typing would be a useful improvement
over optional typing.

Keep in mind that the first (mostly) popular soundly typed language, Java,
came about over two decades after the discovery of type soundness and its
unsound competitor (C++).  In the same way, the creators of Scala or
Haskell continue to believe that they can improve their languages to the
point where they take over vastly more popular and untyped languages such 
as Ruby and Python.

### 2018-01-27 gf-pldi-2018/RESPONSE-PLDI-2018.txt

> There is perhaps another contribution: the idea of "locally-defensive"
> embedding.  Frustratingly, I can't figure out if this is an original
> contribution of the paper, or if it's taken from somewhere else.

The locally-defensive embedding is a mathematical characterization of
transient semantics, and our locally-defensive Racket implements it.

In contrast, the POPL 2017 paper by Vitousek et al. defines transient by
its implementation, which does not explain the properties that an
implementation needs to satisfy in order to be called "transient". So it is
unclear, for example, how to give a transient semantics for the
gradually-typed lambda calculus or a language that is not Python. Our work
fills this gap.

Vitousek [POPL 2017] does not compare type soundness and tag soundness
--------------------------------------------------------------------------------

@@

#99B says

> The main result is a confirmation that the transient semantics of Vitousek
> (POPL 2017) is indeed better-performing than the usual semantics with eager
> casts. This is hardly surprising given the experiments in that paper.

Review #99B sells our results short:

0. Vitousek et al. do _not_ report the performance of programs that mix typed
   and untyped code. (They report the performance of fully typed RP
   programs over fully untyped RP programs; they don't even report the
   slowdown over Python.)

1. Our paper provides the first controlled comparison of type soundness and
   type-tag soundness on the _same_ language platform on mixed-typed code.

2. Our paper provides a mathematical comparison between type soundness and tag
   soundness (Vitousek et al. conflate the two; see their Corollary 5.5.1).

@@

The idea of a spectrum has NOT been previously explored
--------------------------------------------------------------------------------

#99B says:

> The other claimed contribution is the idea that there is in fact a spectrum
> of soundness and performance: but this has been previously explored in the
> work on like types, at least.

Review #99B mischaracterizes like types. The like type approach describes a
_novel type system_ with a **conventional** notion of type soundness. (BTW,
it is also restricted to first--order settings.)

### 2018-01-28 gf-pldi-2018/RESPONSE-PLDI-2018.txt

We need to rewrite this sentence.

The idea is that most languages with runtime type-tags, such as OCaml, normally
have a tag to say "these bits represent an integer" and no tag to say "these
bits represent a natural number".  Therefore, the runtime type-tags in these
languages must be generalized in order to support a gradual type system that
distinguishes Int and Nat types.

> line 380 "...definition of LM  does not include a semantics".  What does it
> mean to "include a semantics".  By "a semantics" I think you mean a definition
> of R and R'.

By "include a semantics", we mean "to define a notion of reduction R and a
reduction relation `e -->_R^* A`.  In contrast, figures 2 3 5 6 7 8 and 9 all
"include a semantics".

@@

> (Incidentally, why use R and R' for static and dynamic rather than R^S and
> R^D?  I kept stumbling over which was which.)

This is a good suggestion.

> Line 284. I have not clue what the bow-tie thing is doing.  It did not seem
> to be used subsequently.

Figures 6 7 8 and 9 use the bow-tie to combine two notions of reduction (one for
statically-typed expressions, one for dynamically-typed expressions) into two
reduction relations (one for static, one for dynamic) that are defined for
expressions that mix statically-typed and dynamically-typed code.

### 2018-05-02 gf-icfp-2018/RESPONSE-ICFP-2018.txt

ICFP '18 Paper #21 Author Response
===========================================================================
Paper #21 A Spectrum of Soundness and Performance

General Response
===========================================================================

- Thank you all for the comments.
- Lots of places we can clarify related work.
  - stw PLDI 2015
- Review B raises serious questions 
  * Kuhlenschmidt
  * Chung
  * Vitousek

Response to Review #21A
===========================================================================

> A weak part of the paper---which is easy to fix---is that there is no
> discussion of mixed-semantics languages.  As such, Dart is badly characterized
> in this paper as an optional/erasure style language.

We will improve the discussion of Dart.
Mixed-semantics is where we plan to go next with this work.

TODO

> Additionally, because the paper mentions universal types, it seems necessary to
> explicitly clarify (maybe as a footnote) that the discussion of type soundness
> elaborated in this work is syntactic; for migratory/gradual typing to ensure (a
> form of) relational parametricity for universal types is another level of
> complexity, and a very active field of research.

Yes, definitely.

> - In Figure 4, there is quite some redundancy between ->^*_{S-N} and
>   ->^*_{D-N}: would it be possible to further factor out the common parts?

We tried factoring the common parts into a meta-function, but decided it was
better to be simple and clear.

> - Please explain why errors do not propagate through E^\bullet contexts

Errors do propagate through ---- E^\bullet is a subset of E.

> - In Theorem 2.0, 2nd reduction alternative, I don't understand why the right
>   hand side cannot be simply: E[dyn t' E* [TagErr]], considering that e' steps
>   to TarErr and that the reduction relation admits reduction under E*.
> (same question for Theorem 2.6)

YES THANK YOU, much simpler.

TODO double-check

> - (l338) "A similar lemma does not hold": would help to state the lemma that
> does not hold (there is space).

Thank you, we will elaborate.

> - The data of Figure 12 could be plotted (again, there is space)

Good point. This table is easy to miss!

> - The discussion of the gradual guarantee in section 5 is a bit incomplete.

Thank you, we will fix the discussion.

Response to Review #21B
===========================================================================

> + The paper does a thorough and controlled performance experiment that
>   confirms that the results of Vitousek et al. (2017) on Reticulated
>   Python carry over to the setting of Typed Racket. That is, the
>   transient semantics avoids the catastrophic performance problems
>   that occur in naive implementations of the natural approach, but
>   transient incurs overhead in statically typed code.

Remarks like this are exactly why we prepared this paper. We are tired of
seeing claims that make generalizations over different languages with
different type systems and different guarantees.

> * The semantic framework is a straightforward adaptation of
>   the Matthews-Findler multi-language semantics.

The first sentence of Section 2.2 and first paragraph of Section 6
(the conclusion) acknowledge Matthews & Findler. If this is not enough
acknowledgement, let us know and we can add more.

> * Theorem 2.0 is a just a combination of the usual Type Safety theorem
>   for gradually typed languages (Siek and Taha 2006) and the Blame
>   Theorem (Tobin-Hochstadt and Felleisen 2006, Wadler and Findler
>   2009). The paper does not discuss these relationships.

The novelty of the theorems is that it shows how 2 approaches to migratory
typing lead

1. we will clarify
2. \vdash different from \vdash_N ! if you don't expose proxies, surface vs. core
3. soundness for a pair

TODO double-check

> * Theorems 2.1, 2.3, and 2.4 are all the usual type safety theorem for
>   dynamically typed languages.

Point 2 still holds --- proxies in the core language.

But right the point of these is not something novel, but to contrast

You are right that all this is _implied_ by the earlier work on gradual typing.
But somehow over the past 10 years we moved from implications to a "type soundness"
theorem published in Vitousek et al. 2017 that is fundamentally different from
a standard type soundness theorem. What happened? I don't know but I think
its a result of glossing over details like this.

TODO

> * Theorem 2.5 is just type safety for a statically typed program, and
>   is a corollary of the equivalence between gradual typing and static
>   typing on fully annotated programs (Siek et al. 2015).  The paper
>   does not discuss this close relationship.

Thank you, we will discuss the relationship.

(This may be a good point in the paper to qualify our remark in the introduction
 about soundness for a single language vs. soundness for a pair.)

> * Theorem 2.6 is an adaptation of Corollary 5.5.1 (Type soundness) of
>   Vitousek et al. (2017).  The paper does not discuss this close
>   relationship.

No, Theorem 2.6 is not just an adaptation of Vitousek et al.'s "type soundness".
Their Corollary 5.5.1 is about a single language.

You might argue that our Theorem 2.6 is just an adaptation of Vitousek et al.'s
Theorem 5.5 (Open-World Soundness), but there are three important differences:

- Vitousek et al. do not define a type system for the surface language;
  it is impossible to reason about programs independent of their core-language
  translation (via Vitousek's \rightsquigarrow)

- Vitousek et al.'s "stuck terms" are defined in terms of the core language;
  it is not clear that evaluation can only get stuck in the context.

> * Theorem 2.7 should be another example of the usual type safety
>   theorem for a dynamically typed language, but it looks like there
>   is a typo in saying |-_{LD} v : \lfloor \tau \rfloor
>   because there is no \tau to refer to here.

Thank you, that is definitely a typo.

> p1. The abstract is much too short.

The abstract says just what we choose it to say --- nothing more.

Please let us know if there is something that NEEDS to be in the abstract and
cannot wait until the introduction.

> p1. "But, it may impose a huge run-time cost"
> It would be appropriate to cite recent work by Kuhlenschmidt et
> al. (see above) that shows that the natural semantics does not have to
> impose a huge run-time cost.

We disagree, it would not be appropriate to cite Kuhlenschidt et al. here.
The sentence you quote is trying to communicate the **historical fact** that
the performance of the natural embedding has led researchers and implementors
to consider other alternatives.

Kuhlendschmidt et al.'s work does not change the past.

> p1. In the discussion of "type soundness", this paper should discuss
> the Blame Theorem, which provides the same guarantees that are provided
> by the theorems in this paper.

All the blame theorems we know* are guarantees about casts in an intermediate
language. Our type soundness theorems are guarantees about the results of
evaluating a surface-language term. These are different guarantees.

`*` Wadler and Findler (2009), Ahmed et al. (2011), Vitousek et al. (2017) ---
    all of which are slightly different theorems!

> p9. "Second, monitoring adds prohibitive run-time cost."
> That is only true of naive implementations. Again, see
> Kuhlenschmidt et al.

Kuhlenschmidt et al's work is not relevant here --- the sentence you quote is
summarizing the motivation for transient given in Vitousek et al. (2017).

Please let us know if you believe we have mis-represented Vitousek et al.

> p4. It is confusing to use boundary errors for both cast errors and
> for errors such as div-by-zero.

We do this to show that soundness for a pair of languages is a generalization
 of soundness for a single "Landin-style" language with some \delta functions.

> p9. "near-constant time"
> Why only "near" and not just "constant time"?

One reason is for union types. The check for `(U String Integer)` must do
at most two constructor checks.

Vitousek et. al 2017 do a near-constant time check for objects; they check
 that the incoming value binds the right fields.

TODO clarify, improve

> p10.
> Why do you perform checking of parameters in the reduction rules
> but insert all the other checks during compilation? This seems
> rather inconsistent and differs, as you later point out,
> from the implementations.

Footnote 3 (line 537) explains.

> p16.
> "This table demonstrates that for pathological examples..."
> Those examples are not pathological.

If you suggest a better word than "pathological" to describe "examples
with worst-case performance", we will consider using it.

> p17. "the implicit conjectures of the literature"
> You seem to be overlooking the performance experiments
> reported on by Vitousek et al. (2017).

Vitousek et al. (2017) do not measure the performance of programs that mix
typed and untyped code.

> p19. "The natural embedding incurs three significantly larger kinds of
> costs." You're forgetting the fourth: the natural embedding also has
> the space-efficiency problem pointed out by Herman et al. (2006),
> which also induces problems with time efficiency.

We did not forget, but you are right we need to be clearer that the cost
of allocation is unbounded in principle. (We will elaborate in Section 4.5.)

> p19. "Sound static types can eliminate the need to dispatch, thus the
> natural embedding's performance can exceed that of the erasure
> embedding"
>
> This paper does not provide empirical results regarding this claim.

See Figure 12.

Response to Review #21C
===========================================================================

> I am a little bit surprised at the fact that static typing does not
> contribute the runtime performance at all.  I would like to see the
> runtime performance on a JIT compiler or a compiler equipped with
> type-conscious optimization.

Typed Racket does do some type-conscious optimization, and Figure 12 shows
that this optimization can make a fully-typed program run faster than a
type-erased program.

Agreed though, it would be very interesting to see similar results for a
language with a type-aware JIT compiler.

### 2018-05-02 gf-icfp-2018/RESPONSE-ICFP-2018.txt

ICFP '18 Paper #21 Author Response
===========================================================================
Paper #21 A Spectrum of Soundness and Performance

General Response
===========================================================================

Thank you for the many comments and suggestions, especially about where our
discussions of related work and future work must improve. See details below.

Review #21B raises serious concerns about works that we failed to cite; namely:

* _A Framework for Object-Oriented Gradual Typing_.
  Chung, Li, Zappa Nardelli, Vitek.
  ECOOP 2018

* _Efficient Gradual Typing_.
  Kuhlenschmidt, Almahallawi, Siek.
  https://arxiv.org/abs/1802.06375

We are familiar with both these works, as Siek and Vitek are co-PIs on a grant
that our second author is PI for (NSF grant SHF 1518844).

For the final version of this paper, we will compare our work to Chung et al.
The main difference between their work and ours is that they do not prove
soundness for their surface languages.

We will not cite Kuhlenschmidt et al. because we are under no obligation to
cite an arXiv paper and (more importantly) we find their performance evaluation
lacking:

- The evaluation for partially-typed programs reports performance relative to
  Racket. This is an apples-to-oranges comparison. As the authors say in
  Section 5.2, "[Grift] does not support a full language".

- The evaluation uses an ad-hoc sampling method, rather than the systematic
  (albiet coarse-grained) methods proposed by Takikawa et al. (2016) and
  Greenman and Migeed (2018). We are not convinced that the Grift sampling
  method supports the strong claims in the paper.

- There are only 6 benchmark programs with data for partially-typed
  configurations; these benchmarks do not include any of the Racket programs
  from Takikawa et al. (2016) or Bauman et al. (2017).

- The data for this paper is not avaiable; it is impossible to tell whether
  the reported running times are on the order of minutes, seconds, or
  milliseconds. (Based on our PI meetings, we suspect it is milliseconds.)

P.S. we will forward this criticism to Kuhlenschmidt et al.

@@

Response to Review #21A
===========================================================================

> A weak part of the paper---which is easy to fix---is that there is no
> discussion of mixed-semantics languages.  As such, Dart is badly characterized
> in this paper as an optional/erasure style language.

We will improve the discussion about Dart and draw attention to mixed-semantics
languages as an area for future work. (It's an area we are excited about.)

@@

> - In Figure 4, there is quite some redundancy between ->^*_{S-N} and
>   ->^*_{D-N}: would it be possible to further factor out the common parts?

We tried factoring the common parts into a meta-function, but decided it was
better to be simple and clear.

> - Please explain why errors do not propagate through E^\bullet contexts

Errors do propagate through, because E^\bullet is a subset of E.

@@

> - In Theorem 2.0, 2nd reduction alternative, I don't understand why the right
>   hand side cannot be simply: E[dyn t' E* [TagErr]], considering that e' steps
>   to TarErr and that the reduction relation admits reduction under E*.
> (same question for Theorem 2.6)

You are right, thank you.

(The 2nd alternative is an artifact of an earlier version of the semantics.)

@@

Thanks for pointing this out.

Response to Review #21B
===========================================================================

> + The paper does a thorough and controlled performance experiment that
>   confirms that the results of Vitousek et al. (2017) on Reticulated
>   Python carry over to the setting of Typed Racket. That is, the
>   transient semantics avoids the catastrophic performance problems
>   that occur in naive implementations of the natural approach, but
>   transient incurs overhead in statically typed code.

Remarks like this are exactly why we prepared this paper. It is time to stop
making broad claims about different languages with different type systems and
different guarantess. We're supposed to be scientists, not marketers.

@@

> * The semantic framework is a straightforward adaptation of
>   the Matthews-Findler multi-language semantics.

The first sentence of Section 2.2 and first paragraph of Section 6
(the conclusion) acknowledge Matthews & Findler. If this is not enough,
we can add more.

> * Theorem 2.0 is a just a combination of the usual Type Safety theorem
>   for gradually typed languages (Siek and Taha 2006) and the Blame
>   Theorem (Tobin-Hochstadt and Felleisen 2006, Wadler and Findler
>   2009). The paper does not discuss these relationships.

The paper does not claim that these theorems are novel. The important
contribution is expressing the different approaches in a framework that
draws attention to their meaningful distinctions.

That said, the difference between Theorem 2.0 and prior work is that it
explicitly connects the surface language (without proxy/monitor values) to the
core language (with proxy values). This is an important distinction for
language designers: even if the surface language does not expose proxies,
users need to understand proxies to understand the language's soundness
guarantee.

@@

> * Theorems 2.1, 2.3, and 2.4 are all the usual type safety theorem for
>   dynamically typed languages.

Again, the paper does not claim these theorems are novel but it is definitely
**unusual** that these theorems include 2 typing judgments.

These theorems also emphasize that soundness for a pair of languages comes
as a pair of theorems.

@@

> * Theorem 2.5 is just type safety for a statically typed program, and
>   is a corollary of the equivalence between gradual typing and static
>   typing on fully annotated programs (Siek et al. 2015).  The paper
>   does not discuss this close relationship.

Good point, we will discuss the relationship.

> * Theorem 2.6 is an adaptation of Corollary 5.5.1 (Type soundness) of
>   Vitousek et al. (2017).  The paper does not discuss this close
>   relationship.

Theorem 2.6 is **not** an adaptation of Vitousek et al.'s "type soundness".
Vitousek et al.'s Corollary 5.5.1 is about a single language.

Nor is our Theorem 2.6 an adaptation of Vitousek et al.'s Theorem 5.5
(Open-World Soundness). There are three important differences:

- Vitousek et al. do not define a type system for the surface language;
  it is impossible to reason about programs independent of their core-language
  translation (via Vitousek's $\rightsquigarrow$).

- Vitousek et al.'s "stuck terms" are defined in terms of the core language;
  it is not clear that typed programs cannot get stuck without studying their
  translation into the core language. In contrast, our framework makes it
  clear that only a dynamically-typed subexpression can step to a TagError.

We will add a clarification to the paper.

@@

Thank you, that is definitely a typo.

> p1. The abstract is much too short.

The abstract says just what we choose it to say --- nothing more.

If you believe something is missing from the abstract, we will consider
including it.

@@

We disagree, it would not be appropriate to cite Kuhlenschidt et al. here.
The sentence you quote is trying to communicate the **historical fact** that
the performance of the natural embedding has led researchers and implementors
to consider other alternatives.

Kuhlendschmidt et al.'s work does not change the past.

> p1. In the discussion of "type soundness", this paper should discuss
> the Blame Theorem, which provides the same guarantees that are provided
> by the theorems in this paper.

All the blame theorems we know (see below) are guarantees about casts in an
intermediate language. Our type soundness theorems are guarantees about the
results of evaluating a surface-language term. These are different guarantees.

"The Blame Theorem"s:

- Wadler and Findler (2009)
- Ahmed et al. (2011)
- Vitousek et al. (2017)

@@

Kuhlenschmidt et al's work is not relevant here --- the sentence you quote is
summarizing the motivation for transient given in Vitousek et al. (2017).

Please let us know if you believe we have mis-represented Vitousek et al.

> p4. It is confusing to use boundary errors for both cast errors and
> for errors such as div-by-zero.

We do this to show how soundness for a pair of languages is a natural
generalization of soundness for a single language (i.e., a language that
includes primitive operations such as division in its trusted code base).

We will clarify in the paper.

@@

> p9. "near-constant time"
> Why only "near" and not just "constant time"?

One reason is for union types. The check for `(U String Integer)` must perform
at most two constructor checks at runtime.

Another reason, from Vitousek et al. (2017), is for structural object types.
Checking that an object has the expected members requires one check for each
member in the structural type.

@@

> p10.
> Why do you perform checking of parameters in the reduction rules
> but insert all the other checks during compilation? This seems
> rather inconsistent and differs, as you later point out,
> from the implementations.

Footnote 3 (line 537) explains.

> p16.
> "This table demonstrates that for pathological examples..."
> Those examples are not pathological.

Suggestions welcome. We just want a word to describe "examples with worst-case
performance".

@@

> p17. "the implicit conjectures of the literature"
> You seem to be overlooking the performance experiments
> reported on by Vitousek et al. (2017).

Vitousek et al. (2017) do not measure the performance of programs that mix
typed and untyped code.

> p19. "The natural embedding incurs three significantly larger kinds of
> costs." You're forgetting the fourth: the natural embedding also has
> the space-efficiency problem pointed out by Herman et al. (2006),
> which also induces problems with time efficiency.

We did not forget, but you are right we need to be clearer that the cost
of **allocation** is unbounded in principle. (We will elaborate in Section 4.5.)

### 2018-05-03 gf-icfp-2018/RESPONSE-ICFP-2018.txt

* _A Framework for Object-Oriented Gradual Typing_.
  Chung, Li, Zappa Nardelli, Vitek.
  ECOOP 2018

* _Efficient Gradual Typing_.
  Kuhlenschmidt, Almahallawi, Siek.
  https://arxiv.org/abs/1802.06375

We are familiar with both these works, as Siek and Vitek are co-PIs on a grant
that our second author is PI for (NSF grant SHF 1518844).

For the final version of this paper, we will compare our work to Chung et al.
The main difference between their work and ours is that they do not prove
soundness for their surface languages (only for the core language).

@@

Response to Review #21A
===========================================================================

> A weak part of the paper---which is easy to fix---is that there is no
> discussion of mixed-semantics languages.  As such, Dart is badly characterized
> in this paper as an optional/erasure style language.

We will improve the discussion about Dart and draw attention to mixed-semantics
languages as an area for future work. (It's an area we are excited about.)

> Additionally, because the paper mentions universal types, it seems necessary to
> explicitly clarify (maybe as a footnote) that the discussion of type soundness
> elaborated in this work is syntactic; for migratory/gradual typing to ensure
> (a form of) relational parametricity for universal types is another level of
> complexity, and a very active field of research.

@@

Thanks for pointing this out.

Response to Review #21B
===========================================================================

> + The paper does a thorough and controlled performance experiment that
>   confirms that the results of Vitousek et al. (2017) on Reticulated
>   Python carry over to the setting of Typed Racket. That is, the
>   transient semantics avoids the catastrophic performance problems
>   that occur in naive implementations of the natural approach, but
>   transient incurs overhead in statically typed code.

Remarks like this are exactly why we prepared this paper. It is time to stop
putting broad claims about different languages with different type systems and
different guarantess in the scientific literature.

Vitousek et al. (2017) does not directly compare transient to natural.

@@

That said, the difference between Theorem 2.0 and prior work is that it
explicitly connects the surface language (without proxy/monitor values) to the
core language (with proxy values). This is an important distinction for
language designers: even if the surface language does not expose proxies,
users need to understand proxies to understand the language's soundness
guarantee.

> * Theorems 2.1, 2.3, and 2.4 are all the usual type safety theorem for
>   dynamically typed languages.

Again, the paper does not claim these theorems are novel but it is definitely
**unusual** that these theorems include 2 typing judgments.

These theorems also emphasize that soundness for a pair of languages comes
as a pair of theorems. (Non-experts should be wary of papers that advertise
a "Theorem 2.5"-style result without discussing partially-typed programs.)

@@

The abstract says just what we choose it to say --- nothing more.

If you believe something is missing from the abstract, we will consider
including it.

> p1. "But, it may impose a huge run-time cost"
> It would be appropriate to cite recent work by Kuhlenschmidt et
> al. (see above) that shows that the natural semantics does not have to
> impose a huge run-time cost.

We disagree, it would not be appropriate to cite Kuhlenschidt et al. here.
The sentence you quote is trying to communicate the **historical fact** that
the performance of the natural embedding led researchers and implementors
to consider other alternatives.

@@

> p17. "the implicit conjectures of the literature"
> You seem to be overlooking the performance experiments
> reported on by Vitousek et al. (2017).

Vitousek et al. (2017) do not measure the performance of programs that mix
typed and untyped code.

> p19. "The natural embedding incurs three significantly larger kinds of
> costs." You're forgetting the fourth: the natural embedding also has
> the space-efficiency problem pointed out by Herman et al. (2006),
> which also induces problems with time efficiency.

We did not forget, but you are right we need to be clearer that the cost
of **allocation** is a space+time cost and is unbounded in principle.

We will elaborate in Section 4.5.

### 2018-05-04 gf-icfp-2018/RESPONSE-ICFP-2018.txt

ICFP '18 Paper #21 Author Response
===========================================================================
Paper #21 A Spectrum of Soundness and Performance

General Response
===========================================================================

Thank you for the many comments and suggestions, especially about where our
discussions of related work and future work must improve.

Review #21B raises two serious concerns. First:

> the novelty of the formal development and theorems is overstated

Our main contributions are:

1. explain three approaches to migratory typing in a framework that supports
   scientific comparisons;
2. systematically compare the implications of each approach for programmers
   reasoning about the behavior of well-typed programs; and
3. systematically compare the performance of each approach.

Our primary claim about the formal development is that it accurately represents
the ideas (some explicit, some implicit) from prior work. We will use the
review comments to more clearly spell out the connections.

Second, Review #21B asks for a discussion of two recent works:

> _A Framework for Object-Oriented Gradual Typing_.
>  Chung, Li, Zappa Nardelli, Vitek.
>  ECCOP 2018
>
> _Efficient Gradual Typing_.
> Kuhlenschmidt, Almahallawi, Siek.
> https://arxiv.org/abs/1802.06375

We are familiar with these, as Siek and Vitek are co-PIs on a grant that
supports our work.

We will acknowledge these works; however, we have serious concerns about the
validity of the conclusions reached by Kuhlenschmidt et al. (We will forward
our concerns to the authors.)

@@

Errors do propagate through, because E^\bullet is a subset of E.

> - In Theorem 2.0, 2nd reduction alternative, I don't understand why the right
>   hand side cannot be simply: E[dyn t' E* [TagErr]], considering that e' steps
>   to TarErr and that the reduction relation admits reduction under E*.
> (same question for Theorem 2.6)

You are right, thank you.

(This 2nd alternative is an artifact of an earlier version of the semantics.)

### 2018-05-04 gf-icfp-2018/RESPONSE-ICFP-2018.txt

Response to Review #21B
===========================================================================

> + The paper does a thorough and controlled performance experiment that
>   confirms that the results of Vitousek et al. (2017) on Reticulated
>   Python carry over to the setting of Typed Racket. That is, the
>   transient semantics avoids the catastrophic performance problems
>   that occur in naive implementations of the natural approach, but
>   transient incurs overhead in statically typed code.

Vitousek et al. (2017) does not directly compare transient to natural.

Remarks like this are exactly why we prepared this paper. It is time to stop
polluting the scientific literature with general claims about different
languages with different type systems and different guarantees.

@@

The first sentence of Section 2.2 and first paragraph of Section 6
(the conclusion) acknowledge Matthews & Findler. If this is not enough,
we can add more.

> * Theorem 2.0 is a just a combination of the usual Type Safety theorem
>   for gradually typed languages (Siek and Taha 2006) and the Blame
>   Theorem (Tobin-Hochstadt and Felleisen 2006, Wadler and Findler
>   2009). The paper does not discuss these relationships.

The paper does not claim that these theorems are novel. The important
contribution is expressing the different approaches in a framework that
draws attention to their meaningful distinctions.

That said, the difference between Theorem 2.0 and prior work is that it
explicitly connects the surface language (without proxy/monitor values) to the
core language (with proxy values). This is an important distinction for
language designers: even if the surface language does not expose proxies,
users need to understand proxies and the evaluation type system to understand
the language's soundness guarantee.

@@

> * Theorems 2.1, 2.3, and 2.4 are all the usual type safety theorem for
>   dynamically typed languages.

Again, the paper does not claim these theorems are novel but it is definitely
unusual that these theorems include 2 typing judgments.

@@

Theorem 2.6 is **not** an adaptation of Vitousek et al.'s "type soundness".
Vitousek et al.'s Corollary 5.5.1 is about a single language.

Nor is our Theorem 2.6 an adaptation of Vitousek et al.'s Theorem 5.5
(Open-World Soundness). There are two important differences:

- Vitousek et al. do not define a type system for the surface language;
  it is impossible to reason about programs independent of their core-language
  translation (via Vitousek's $\rightsquigarrow$).

- Vitousek et al.'s "stuck terms" are defined in terms of the core language;
  it is not clear from the theorem that typed programs cannot get stuck without
  studying their translation into the core language. In contrast, our theorem
  shows that only dynamically-typed expressions can reduce to a TagError.

@@

The abstract says just what we choose it to say --- nothing more.

If you believe something is missing from the abstract, we will consider
including it.

> p1. "But, it may impose a huge run-time cost"
> It would be appropriate to cite recent work by Kuhlenschmidt et
> al. (see above) that shows that the natural semantics does not have to
> impose a huge run-time cost.

We disagree, it would not be appropriate to cite Kuhlenschidt et al. here.
The sentence you quote is trying to communicate the historical fact that
the performance of the natural embedding led researchers and implementors
to consider other alternatives.

@@

Kuhlenschmidt et al's work is not relevant here --- the sentence you quote is
summarizing the motivation for transient given in Vitousek et al. (2017).

Please let us know if you believe we have mis-represented Vitousek et al. (2017)

> p4. It is confusing to use boundary errors for both cast errors and
> for errors such as div-by-zero.

We do this to show how soundness for a pair of languages is a natural
generalization of soundness for a single language that includes primitive
operations (such as division) in its trusted code base.
